Disclosing funding sources for open access publication fees: the Open APC

### 2016-06-04 article.Rmd
retooled discussion, enough for today

To address these problems of fragmented spending on publication fees and of in-transparency about what was being paid, some European research funders and research performing institutions have recently begun to disclose their payment for publication as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc_12] and the Austrian Science Fund FWF [@fwf_apc_13]. The not-for-profit company Jisc followed by collecting data from higher-education institutions in the UK [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report what was being spent along with bibliographic information, including title, journal and publisher, and a persistent identifier to the publisher's version. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates and persistent identifiers to the full text [@Neylon_2014; @woodward_2014]. Parts of Jisc's cost data was examined by @Pinfield_2015. Although the average spending on publication fees remained stable across the universities, they found large price variances, as well as a varying number of articles UK universities supported between 2007 - 2014, confirming earlier studies that collected price information from journal websites.

@@

This paper focuses on how much German universities and research organisation spent on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has strongly influenced how universities manage institutional support for publication fees.[^3] Before the DFG started to pay for centrally funded publication fees on a pro rata basis through its "Open-Access Publishing" program in 2011, and similar to the situation described in Canada [@Hampson_2014] or the UK [@PINFIELD_2012], only few central funds existed [@eppelin_2012]. Because the DFG has enforced a set of criteria grantees have to comply with, similar implementations for supporting open access publishing across German universities exists [@Fournier_2013]. These criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 € (excluding value added tax)[^3]. Grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. This includes central invoicing schemes, and related agreements between university libraries and publishers, leading to discounts for the authors [@Fournier_2013].

@@

Prices also varied within single journals. Based on the number of articles paid for, Table 3 illustrates the top ten out of `r my_apc %>% group_by(journal_full_title) %>% summarize(n()) %>% nrow()` journals. Payments to these journals represent `r apc_journal %>% filter(!journal_full_title == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the overall expenditure. In the case of Atmospheric Chemistry and Physics Discussions, the large price range can be explained by the fact that this journal charges per page and also takes the submission's file format into consideration.

@@

In Germany, institutional spending on open access publication fees has increased over the years, confirming the general trend of publication fees in academic publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. With a share of 99%, the majority of open access articles German institutions reported to the Open APC initiative were published in fully open access journals. This presumably reflects the DFG funding policy which excludes the support of articles published in hybrid open access journals. The DFG has been financially supporting the implementation of central publication funds at more than 30 German universities since 2009. However, in reviewing self-reported cost data from funders or countries that also support hybrid open access journals or open access books, we revealed smaller proportions of payments in favour of articles in fully open access journals. Because open access publication fee spending is fragmented, we cannot answer whether German researchers avoided opting for open access when publishing in hybrid journals or used other budgets to pay publication fees required to make their work open access through these kind of journals.

In our study, Crossref thoroughly indexed open access journal articles in the Open APC data-set. We could gather metadata representing publisher and journal titles for 99 % articles, and successfully merged these information with the Open APC cost data. Using metadata from Crossref, therefore, reduces the extensive validation work of bibliographic information provided that the reporting of the DOI along with the expenses is made mandatory. Drawing on Crossref would also increase the comparability of cost data for future negotiations with publishers on open access agreements, and the open access spending between open data initiatives that apply the same reporting standards, as its metadata represent the dynamic landscape of academic publishing in terms of ongoing mergers of publishing houses or name changes.

Another advantage of self-reported data-sets on the article-level to disclose spending on open access publication fees is that they enable researchers and practitioners alike to study in which open access journals researchers from one institution actually publish, and to compare these findings with that of other universities or research organizations. For instance, our study revealed that the size of publication fee spending differed among the institutions with the Max Planck Society accounting for almost 39 % of the overall articles. Many universities and research organization reported remarkably lower number of supported open access articles to the Open APC initiative. Using self-reported data, therefore, contributes to the understanding about how much and to what extent spending on open access publishing varies on the institutional level. This is particularly relevant given the increasingly important role open access publishing plays in recent negotiations between German universities and research organizations forming consortia on the one side and publishers on the other about financing scholarly publishing in future [@allianz].

This study is limited in some respects. One is that we cannot assess whether publishers and journals granted publication fee discounts seeing that the Open APC initiative does not track this kind of information. However, the large price ranges of particular journals suggests that varying pricing schemes were in place. Adding to this complexity, it is likely that some institutions only paid parts of the publication fee. Take for instance the journal Nature Communication. Charges reported in our sample ranged between 2000 €, the DFG price cap, and 4.403 €. Such co-payments that involve several budgets were a proposed strategy to sustain publication funds at German universities [@Fournier_2013]. In another case, one university included its charges for participating in the German SCOAP consortia and presumably divided the sum by the articles published in SCOAP journals. In this national consortia, managed by the German National Library of Science and Technology, payments were not directly made per article. Instead, subscription costs between a participating library and a publisher were reconciled, and the reduction transferred to the consortia to finance publications in SCOAP journals.

@@

Our analysis on how institutional spending per articles was distributed over publishers and journal titles shows that open access publishing is diverse and concentrated at the same time. While we were able to identify 139 individual publishing houses that were supported by the German universities and research organizations, the distribution is highly skewed. 92 % of open access publication fee spending went to ten publishers, confirming a general high concentration of few publishers in current academic publishing. However, our study could not confirm that publications in open access journals owned by traditional publishing houses account for most of the spending on publication fees. Rather, open access publishers such as Public Library of Science (PloS), Copernicus GmbH or MPDI AG rank higher in our study than in the analyses of cost data in the UK.

This study finally confirms the leading role of "mega-journals" in open access publishing, including the multidisciplinary PLOS ONE and the journals New Journal of Physics, Atmospheric Chemistry and Physics Discussions and Frontiers in Psychology, all of which publish contributions from all branches of their respective discipline. In general, an estimated 14 out of more than 10,000 journals registered in DOAJ in 2015 accounted for up to 15–20 % of all articles published in full open access journals [@Bj_rk_2015].

>>>>

To address these problems of fragmented spending on publication fees and of in-transparency about what was being paid, some European research funders and research performing institutions have recently begun to disclose their payment for publication as open data. To our knowledge, the first research funders providing such data were the Wellcome Trust [@wellcome_apc_12] and the Austrian Science Fund FWF [@fwf_apc_13]. The not-for-profit company Jisc followed by collecting data from UK universities [@Lawson_data_2015]. Disclosed as publicly available spreadsheets, these data-sets self-report what was being spent along with bibliographic information, including title, journal and publisher, and a persistent identifier to the publisher's version. Curatorial efforts focused on the disambiguation of publisher and journal titles as well as on detecting duplicates and persistent identifiers to the full text including the Digital Object Identifier (DOI) [@Neylon_2014; @woodward_2014]. Parts of Jisc's cost data was examined by @Pinfield_2015. Although the average spending on publication fees remained stable across the universities, they found large price variances, as well as a varying number of articles UK universities supported between 2007 - 2014, confirming earlier studies that collected price information from journal websites.

@@

This paper focuses on how much German universities and research organisations spent on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has strongly influenced how universities manage institutional support for these charges. Before the DFG started to pay for centrally funded publication fees on a pro rata basis through its "Open-Access Publishing" program in 2011, and similar to the situation described in Canada [@Hampson_2014] or the UK [@PINFIELD_2012], only few central funds existed [@eppelin_2012]. Because the DFG has enforced a set of criteria grantees have to comply with, similar implementations for supporting open access publishing across German universities exists [@Fournier_2013]. These criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 € (excluding value added tax)[^3]. Grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. This includes central invoicing schemes and memberships, among others, agreed between university libraries and publishers that, in many cases, lead to discounts [@Fournier_2013].

@@

Prices also varied within single journals. Based on the number of articles paid for, Table 3 illustrates the top ten out of `r my_apc %>% group_by(journal_full_title) %>% summarize(n()) %>% nrow()` journals. Payments to these journals represent `r apc_journal %>% filter(!journal_full_title == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the overall expenditure. The multidisciplinary PLOS ONE and the journals New Journal of Physics, Atmospheric Chemistry and Physics Discussions and Frontiers in Psychology, all of which publish contributions from all branches of their respective discipline, were also well represented in the Open APC data-set. In general, an estimated 14 out of more than 10,000 journals registered in DOAJ in 2015 accounted for up to 15–20 % of all articles published in full open access journals [@Bj_rk_2015] In the case of Atmospheric Chemistry and Physics Discussions, the large price range can be explained by the fact that this journal charges per page and also takes the submission's file format into consideration.

@@

In Germany, institutional spending on publication fees charged by open access journals has increased over the years. These findings are consistent with the general trend of publication fees as revenue source for open access publishing [@Davis_2011; @Laakso_2012; @Pinfield_Review_2015]. They also reflect the increasing importance of providing central support to cover these charges [@Fournier_2013; @Kingsley_2014]. Similar to investigations of the expenditure on publication fees at an institutional level in the UK [@Pinfield_2015], spending size varies across the German universities and research organisations included in the Open APC initiative. With a proportion of 39 % of the overall number of articles, the Max Planck Society, a large non-university research organisation, supported most open access journal articles. This likely reflects the centralised library support at the Max Planck Society where the Max Planck Digital Library has managed large open access agreements with publishers over the last decade [@Schimmer_2013, @Sikora_2015]. Many universities and research organisations, on the other hand, reported a remarkably lower number of supported articles.

To answer data curation issues while unifying spending data raised by UK initiatives, re-using DOIs to gather bibliographic metadata from Crossref is a promising approach. In our study, Crossref thoroughly indexed open access journal articles disclosed in the Open APC data-set, providing publisher and journal titles for 99 % of the overall article volume. Making use of metadata from Crossref, therefore, reduces the extensive validation work of bibliographic information provided that the DOI is made available in the cost data, a requirement needed to be addressed also in other fields of the quantitative study of scholarly communication [@Haustein_2016]. Drawing on Crossref would also increase the comparability of cost data to prepare future negotiations with publishers on open access agreements, because Crossref's metadata reflects the current landscape of academic publishing in terms of ongoing mergers of publishing houses or name changes. Comparative studies of publication fee spending using data on the article-level would also benefit from such an approach.

This study is limited in some respects. One is that we cannot assess whether publishers and journals granted publication fee discounts because the Open APC initiative does not track this kind of information. However, large price ranges of particular journals suggests that varying pricing schemes are in place, an observation also made earlier [@Solomon_2012, @Pinfield_2015]. Adding to this complexity, it is likely that some institutions only paid parts of the publication fee. Take for instance the journal Nature Communication. Charges reported to the Open APC initiative ranged between 2000 €, the DFG price cap, and 4.403 €. Although such payments made from several budgets are a proposed strategy to sustain publication funds at German universities [@Fournier_2013], these pro rata amounts cannot be recognised as such from the Open APC data, leading to a possibly flawed representations of publication fee spending in Germany. In another case, one university included its charges for participating in the German SCOAP consortia and presumably divided the sum by the articles published in SCOAP journals. In German SCOAP consortia, managed by the German National Library of Science and Technology, payments were not directly made per article. Instead, subscription costs between a participating library and a publisher were reconciled, and the reduction transferred to the SCOAP to finance publications in particle physics journals.

@@

Our analysis on how institutional spending per articles was distributed over publishers and journal titles indicates that open access publishing is heterogeneous and concentrated at the same time. While we were able to identify 139 individual publishing houses that were supported by the German universities and research organizations, the distribution is highly skewed. 92 % of open access publication fee spending went to ten publishers, being consistent with a observed high concentration of few publishers in current academic publishing [@Larivi_re_2015]. However, our study could not confirm that publications in open access journals owned by traditional publishing houses account for most of the spending on publication fees as observed by @Pinfield_2015. Rather, open access publishers such as Public Library of Science (PloS), Copernicus GmbH or MPDI AG rank higher in our study than in the analyses of cost data in the UK.

One possible explanation why traditional publishers are less well represented in our study, is the lack of cost information about hybrid open access journals. 99 % of the overall article volume German universities and research organisations financially supported were published in fully open access journals. This presumably reflects the DFG funding policy which excludes the support of articles published in hybrid open access journals through many university open access funds. However, reviewing self-reported cost data from funders or countries that also support hybrid open access journals or open access books, we observe smaller proportions of payments in favour of articles in fully open access journals. Because open access publication fee spending is fragmented, it remains open to speculation whether authors affiliated with universities and research organisations simply avoided opting for open access when publishing in hybrid journals for political or economical reasons, or used other budgets to pay publication fees required to make their work open access through these kind of journals.
